---
title: "Day1. Making Computational Methods Accessible: R Basics and Data Wrangling"
author: "amber_du"
output: 
  html_document
---

# R and RStudio  
## What is R and what is RStudio?  
### R
R is a programming language for statistical computing and graphics. You can consider it as an environment where techniques are implemented and it can be further extended via packages. You can download R at [CRAN (the comprehensive R archive network)](https://cran.rstudio.com).  
### RStudio 
RStudio is an IDE(Integrated Development Environment) for R. You can download RStudio at [this link](http://www.rstudio.com/download).  

# Getting Started 
## R Script

## R Markdown 

## Operators 

### Assignment Operators 
Assignment operators include the left assignment operators: <-, =, <<- and right assignment operators: -> and ->>. In general, <- and = are not the same: <- can be used anywhere and the use of = is restricted. 
```{r assignment-operators}
a <- c(1,2,3) # numeric vector; c() [concatenation] is used to create a vector 
a # view object a 
c(TRUE, FALSE) -> b
b
c <- c("a","b") # character vector
c
c(FALSE, 1) ->> d
d
```
### Arithmetic Operators
Arithmetic operators includes + (add), - (subtract), * (multiply), / (divide), %% (remainder), ^ (power), ... 
```{r arithmetic-operators}
a <- 2**3
a
(b<-2^3) # assign 2*3 to b and view the result 
c(1,2)-2
c(1,2,3,4,5,6)-c(1,2) # longer object length is a multiple of shorter length; vector recycling: will talk about this later 
c(1,2) * c(2,3,4) # longer object length is not a multiple of shorter length 
c(4,6,7) %%2
```

### Relational Operators 
Relational operators includes > (greater than), < (less than), == (equal), <= (less than or equal to), >= (greater than or equal to), != (unequal)
```{r relational-operators}
TRUE==1;FALSE==0 # use of semicolon for multiple commands on one line 
c(1,2,3,4) <= c(2,3,4,5) 
c(1,2,3,4) != c(2,3,4,5) 
```
### Miscellaneous Operators 
```{r miscellaneous-operators}
1:20 # a series of number in sequence 
seq(1,20, by=2)
rep(3,2) # repeat 3 twice 
rep("repeat",5)
7 %in% c(1,2,3,4,5)
```
## Work with objects 
```{r work-with-objects, eval=FALSE}
ls() # list all the objects 
rm(a) # remove object "a"
rm(list=ls()) # clean up the environment 
help(exp) # get help; this is the same as ?exp
example(exp)
getwd() # get working directory 
setwd()  # set working directory
```

## Functions 
### Built-in Functions 
**Math functions**
```{r math-functions}
a <- c(-1,2)
abs(a) # the absolute value of input "a"
sqrt(4) # the square root of input "a"
log(10) # the natural log of input "a"
log10(10) # the common log of input "a"
```
**String functions** 
```{r stirng-unctions}
paste('one',2,'three', sep=', ') # concatenate vectors after converting to character 
strsplit("This is a line.", " ") # split the elements of a character vector into substrings
```
**Statistical functions**
Some useful statistical functions includes max(), min(), sum(), ...
```{r statistical-functions}
(r.x <- range(stats::rnorm(100)))
```
### User-defined functions  
There are three steps in defining a function: its name, inputs and body. Writing a function has a few advantages compared with copy-and-paste. For example, as the requirement / goal changes, you only need to make update in the function, instead of in multiple places. 
```{r user-defined-functions} 
sum_v2 <- function(a,b) { # argument are not stored in Environment 
  return(sum(a:b))
}
# stored under Functions in Environment 
sum_v2(1,10)
```
## Packages in R 
An R package is a collection of functions, data, and documentation that extends the capabilities of R. 
```{r install-and-load-ggplot2, eval=FALSE}
install.packages("ggplot2") # install 
library(ggplot2) # load ggplot2 in Environment 

pkgs <- c("ggplot2", "purrr")
install.packages(pkgs)
```
## Vectors in R 
### Atomic vectors (homogeneous)
**Logical**
**Numeric: Integer, Double**
**Character**
### Lists (heterogeneous)
### Augmented vectors 
```{r vectors-in-R}
# atomic vectors 
a <- c(1:10, NA) # 1 special values in a integer vector 
b <- c("a", "b", "c")
c <- c(1.2, 1.0, 2.22, NA, NaN, Inf, -Inf) # 4 special values in a double vector 
is.na(NA); is.na(NaN)
is.nan(NA); is.nan(NaN)
d <- c(TRUE, FALSE, NA) # 3 possible values in a logical vector 
typeof(a)
typeof(b);length(b)
typeof(c)
typeof(d)

# doubles
sqrt(2) == 1.414214
dplyr::near(sqrt(2), 1.414214, tol=0.001)

#list 
e <- list(1,2, "a", 1:10)
length(e)
list(a, b, c)

# matrix
matrix <- matrix(c(1,2,3,4,5,6,7,8), nrow=2)
# dataframe 
df <- data.frame(
  "x" = 1:3,
  "y" =  b, 
  "z" = c(1, "a", "b")
)

```
### Coercion
```{r coercion}
# explicit 
as.logical("TRUE"); as.logical(0); as.logical(1); as.logical(3)
as.integer(1.2); as.integer(1.5); as.integer(1.9)

# implicit
sum(c(TRUE, FALSE, FALSE))
mean(c(TRUE, FALSE, FALSE))

set.seed(715)
x <- rnorm(100)
y <- x > 1.96 |# which one in x is greater than 1.96
mean(y) # what proportion in x is greater than 1.96

typeof(1); typeof(1L)
typeof(c(TRUE, 2L))
typeof(c(2L, 3.8))
typeof(c(3.8, "x")) # the most complex type wins 
```

### Check the type of vectors
```{r check-type-of-vectors}
double <- c(1,2,3)
typeof(double)
purrr::is_double(double)
```

### Vector Recycling
```{r vector-recycling}
rchisq(100, df=2) > 1
```

### Vector names 
```{r}
c(xx=1, yy=2, zz<-3)
```
### Subsetting vectors 
```{r}
chr <- c("a", "b", "c")
chr[c(1,2)]
chr[c(1,2,1,2)]
chr[c(-1)] # do not select the value at position 1

int <- c(1,2,3,4,5)
int[int>1] # select the values that corresponds to TRUE

int <- c(xx=1, yy=2, zz=3)
int[c("xx")]

matrix[1,]
matrix[,1]
# filter in a data.frame 
dplyr::filter() 
```
## List
```{r list}
complicated_list <- list(1:2, c(1,2), "a", "b", c("a", "this is b"), c(1, "a", TRUE))
str(complicated_list)

tr <- list(c(1,2), c(3,4))
str(tr)
tr_2 <- list(list(1,2), list(3,4))
str(tr_2)
```
# Getting Help





