---
title: 'Day1. Making Computational Methods Accessible: R Basics and Data Wrangling'
author: "Amber Du"
date: "July 15, 2023"
output:
  html_document: default
  pdf_document: default
---

# R and RStudio  
## What is R and what is RStudio?  
### R
R is a programming language for statistical computing and graphics. You can consider it as an environment where techniques are implemented and it can be further extended via packages. You can download R at [CRAN (the comprehensive R archive network)](https://cran.rstudio.com).   

### RStudio 
RStudio is an IDE(Integrated Development Environment) for R. You can download RStudio at [this link](http://www.rstudio.com/download).  

### Command line tool in R 
```{r command-line-tool-in-r, eval=FALSE}
cd
mkdir 
pwd
echo
head 
more 
```

# Getting Started 
**R Script**  
**R Markdown**   

## Operators   

### Assignment Operators 
Assignment operators include the left assignment operators: `<-`, `=`, `<<-` and right assignment operators: `->` and `->>`. In general, `<-` and `=` are not the same: `<-` can be used anywhere and the use of `=` is restricted. We will talk about their differences later, but `<-` is recommended. 
```{r assignment-operators}
a <- c(1,2,3) # numeric vector; c() [concatenation] is used to create a vector 
a # view object a; call variable a 
c(TRUE, FALSE) -> b
b
c <- c("a","b") # character vector
c
c(FALSE, 1) ->> d
d
```
### Arithmetic Operators
Arithmetic operators includes `+` (add), `-` (subtract), `*` (multiply), `/` (divide), `%%` (remainder), `^` (power), ... 
```{r arithmetic-operators}
a <- 2**3
a
(b<-2^3) # assign 2*3 to b and view the result 

c(1,2)-2
c(1,2,3,4,5,6)-c(1,2) # longer object length is a multiple of shorter length; vector recycling: will talk about this later 
c(1,2) * c(2,3,4) # longer object length is not a multiple of shorter length 
c(4,6,7) %%2
```

### Relational Operators 
Relational operators includes `>` (greater than), `<` (less than), `==` (equal), `<=` (less than or equal to), `>=` (greater than or equal to), `!=` (unequal)
```{r relational-operators}
TRUE==1;FALSE==0 # use of semicolon for multiple commands on one line 
c(1,2,3,4) <= c(2,3,4,5) 
c(1,2,3,4) != c(2,3,4,5) 
```
### Miscellaneous Operators 
```{r miscellaneous-operators}
1:20 # a series of number in sequence 
seq(1,20, by=2)
rep(3,2) # repeat 3 twice 
rep("repeat",5)
7 %in% c(1,2,3,4,5)
```
## Working with R objects 
```{r work-with-objects, eval=FALSE}
ls() # list all the objects 
rm(a) # remove object "a"
rm(list=ls()) # clean up the environment 
help(exp) # get help; this is the same as ?exp
example(exp)
getwd() # get working directory 
setwd()  # set working directory
```

## Functions 
### Built-in Functions 
**Math functions**
```{r math-functions}
a <- c(-1,2)
abs(a) # the absolute value of input "a"
sqrt(4) # the square root of input "a"
log(10) # the natural log of input "a"
log10(10) # the common log of input "a"
```
**String functions** 
```{r stirng-unctions}
paste('one',2,'three', sep=', ') # concatenate vectors after converting to character 
strsplit("This is a line.", " ") # split the elements of a character vector into substrings
```
**Statistical functions**
Some useful statistical functions includes `max()`, `min()`, `sum()`, ...
```{r statistical-functions}
(r.x <- range(stats::rnorm(100)))
```
### User-defined functions  
There are three steps in defining a function: its name, inputs and body. Writing a function has a few advantages compared with copy-and-paste. For example, as the requirement / goal changes, you only need to make update in the function, instead of in multiple places. 
```{r user-defined-functions} 
sum_v2 <- function(a,b) { # argument are not stored in Environment 
  return(sum(a:b))
}
# stored under Functions in Environment 
sum_v2(1,10)
```
## Packages in R 
An R package is a collection of functions, data, and documentation that extends the capabilities of R. 
```{r install-and-load-ggplot2, eval=FALSE}
install.packages("ggplot2") # install 
library(ggplot2) # load ggplot2 in Environment 

packageVersion("ggplot2") # check the package version 
packageVersion("dplyr")
# update.packages("dplyr") update the package 

pkgs <- c("ggplot2", "purrr")
install.packages(pkgs)
```
## Vectors and augmented vectors in R  
The section below for vectors and augmented vectors follows [Section 20 from R for Data Science](https://r4ds.had.co.nz/vectors.html?q=matrix#vectors)
**Atomic vectors (homogeneous)**  
Logical   
Numeric: Integer, Double   
Character    
**Lists (heterogeneous)**   
**Augmented vectors**   
Factors   
Dates and date-times   
Tibbles   
### Atomic vectors  
```{r vectors-in-R}
# atomic vectors 
a <- c(1:10, NA) # 1 special values in a integer vector 
b <- c("a", "b", "c")
c <- c(1.2, 1.0, 2.22, NA, NaN, Inf, -Inf) # 4 special values in a double vector 
is.na(NA); is.na(NaN)
is.nan(NA); is.nan(NaN)
d <- c(TRUE, FALSE, NA) # 3 possible values in a logical vector 
typeof(a); class(a)
typeof(b);length(b); class(b)
typeof(c); class(c)
typeof(d); class(d)


# doubles
sqrt(2) == 1.414214
dplyr::near(sqrt(2), 1.414214, tol=0.001)
```
#### Add elements to a vector 
```{r add-elements-to-a-vector}
(a <- c(1:10, NA))
(a <- c(a, 11))
```
#### Structure of atomic vector 
```{r structure-of-atomic-vector}
c(1, c(2, c(3, 4))) == c(1,2,3,4)
```
#### Coercion
```{r vector-coercion}
# explicit 
as.logical("TRUE"); as.logical(0); as.logical(1); as.logical(3)
as.integer(1.2); as.integer(1.5); as.integer(1.9)

# implicit
sum(c(TRUE, FALSE, FALSE)) # logical vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0
mean(c(TRUE, FALSE, FALSE))
1 < "2"

set.seed(715)
x <- rnorm(100)
y <- x > 1.96 # which one in x is greater than 1.96
mean(y) # what proportion in x is greater than 1.96

typeof(1); typeof(1L)
typeof(c(TRUE, 2L))
typeof(c(2L, 3.8))
typeof(c(3.8, "x")) # the most complex type wins: types from least to most complex are: logical, integer, double, and character. 
```

#### Check the type of vectors
```{r check-type-of-vectors}
double <- c(1,2,3)
typeof(double)
purrr::is_double(double)
```

#### Vector Recycling
```{r vector-recycling}
rchisq(100, df=2) > 1
```

#### Vector names 
```{r vector-names}
c(xx=1, yy=2, zz<-3) # compare with c(xx=1, yy=2, zz=3)

x <- c(1,2,3)
names(x)

y <- c(xx=1,yy=2,zz=3)
names(y)
```
#### Subsetting vectors  
Index in R starts from 0 (vs. index in python starts from 0)  
```{r}
chr <- c("a", "b", "c")
chr[1] # chr[4]
chr[c(1,2)]
chr[c(1,2,1,2)]
chr[c(-1)] # do not select the value at position 1

int <- c(1,2,3,4,5)
int[int>1] # select the values that corresponds to TRUE

int <- c(xx=1, yy=2, zz=3)
int[c("xx")]
```


### List
```{r list}
#list 
f <- list(1,2, "a", 1:10)
length(f)
list(a, b, c)
```

#### Structure of a list    
```{r structure-of-a-list}
# structure of a list 
complicated_list <- list(1:2, c(1,2), "a", "b", c("a", "this is b"), zzz=c(xxx=1, "a", TRUE))
str(complicated_list)

tr <- list(c(1,2), c(3,4))
str(tr)
tr_2 <- list(list(1,2), list(3,4))
str(tr_2)
```
#### Coercion 
```{r coercion}
int <- c(1,2,3,4,5,6)
ls <- as.list(int)
str(ls)

int_back <- unlist(ls) # turn a list into an atomic vector
str(int_back)
```
#### Subsetting list 
```{r subsetting-list}
# subsetting list 
str(complicated_list[1:3]) # still a list 
str(complicated_list[6])
str(complicated_list[[6]]); str(complicated_list$zzz) # remove one level of hierarchy; these two are equivalent 
str(complicated_list[[6]][1]); str(complicated_list[[6]][[1]])
```
#### Examples of list 
```{r example-of-list}
purrr::is_list(lm(mtcars$mpg ~ mtcars$cyl)) # linear models are lists 
```
### Factors 
```{r factor}
# factors 
e <- factor(b, levels=c("a", "b", "c", "d")) # factor is a special type of vector; usually used in categorical data analysis 
typeof(e); class(e) # integer; factor 
str(e)
attributes(e)
```

### Matrix  
```{r matrix}
mat <- matrix(c(1,2,3,4,5,6,7,8), nrow=2)
mat
typeof(mat)
attributes(mat)
t(mat)
```
#### Subsetting matrix 
```{r subsetting matrix}
mat[1,]
mat[,1]
```
### Dates
```{r dates}
dat_time <- as.Date("1971-01-01")
typeof(dat_time)
attributes(dat_time)
```
### Data Frame  
```{r dataframe}
df <- data.frame(x = 1:5, y = 5:1)
typeof(df)
attributes(df)

df_2 <- data.frame(
  id = c("A", "B", "C"),
  gender = factor(c("F", "M", "F")),
  grade = c(60, 70, 70)
)
typeof(df_2)
class(df_2)
str(df_2) # show the structure: 3 observations of 2 variables 
```
#### Subsetting data frame  
In R, `&` represent **AND**, `|` represents **OR**
```{r subsetting-dataframe}
class(df_2$grade) # use $ to extract single column; equivalent to df_2[[2]]
typeof(df_2$grade)
mean(df_2$grade)
hist(df_2$grade)

class(df_2[2]) 
typeof(df_2[2])
# mean(df_2[2])
# hist(df_2[2])

class(df_2[c(1,2)]) # extract multiple columns 
typeof(df_2[c(1,2)])

class(df_2[,c(1,2)]) # extract multiple columns
typeof(df_2[,c(1,2)])

class(df_2[,1:2]) # extract multiple columns
typeof(df_2[,1:2])

class(df_2[c("id", "grade")]) # extract multiple columns
typeof(df_2[c("id", "grade")])

class(df_2[1,2]) # extract value from a single cell [rows, columns]
typeof(df_2[1,2])

class(df_2[1,]) # extract row [rows, ]
typeof(df_2[1,])

class(df_2[,2]) # extract column [, column]
typeof(df_2[,2])

df_2[c(1,3), c(1,2)] # select the 1st & 3rd row, 1st and second column 

df_2[df_2$grade == 70, ] # select the rows where grade is 70

df_2[df_2$grade == 70 & df_2$gender == "M", ] # select the rows where grade is 70 and gender is M
# run filter from dplyr package 
# dplyr::filter() 
```

### Tibbles  
```{r tibbles}
tib <- tibble::tibble(x = 1:5, y = 5:1)
typeof(tib)
attributes(tib)
```

## Import and Export 
```{r import-and-export}
# Apply Mobility.csv file is download from https://sicss.io/boot_camp/ under visualization 
apple_mobility <- read.csv("data/apple_mobility_data.csv") # default: header = TRUE, sep = ",",
head(apple_mobility)[, c(1,2,3,4,5)]
str(apple_mobility[, c(1,2,3,4,5)])

library("writexl") 
write_xlsx(df_2, "df_2.xlsx",col_names = TRUE)
write.csv(df_2, "df_2.csv")
```
# Getting Help





